---
title: "Non-standard Applications of ksformat for Clinical Trials and Other Tasks"
format:
  html:
    toc: true
    number-sections: true
execute:
  enabled: false
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ksformat)
```

# Overview

`ksformat` is usually described as a SAS PROC FORMAT-like tool for ordinary
value-to-label lookup. In practice, the package is useful in places where a
clinical trial workflow needs more than a plain dictionary:

- protocol-defined visit windows
- subject- or arm-specific bucketing rules
- composite keys built from multiple columns
- reverse lookups for data cleaning and QC
- report-ready numeric display for listings and summaries

This article shows a few non-standard uses that come up in clinical trials and
adjacent data tasks.

# 1. Protocol windows and subject-specific buckets

A common clinical-trials task is mapping visit-day values into protocol-defined
windows. The same idea extends to arm-specific rules, where one stratum has a
slightly different window structure from another.

```{r}
#| label: protocol-windows
windows <- fmap_strata(
  stratum = c("ARM_A", "ARM_A", "ARM_A", "ARM_B", "ARM_B"),
  low     = c(0,        7,       28,      0,       14),
  high    = c(7,        28,      Inf,     14,      Inf),
  label   = c("Baseline", "Wk1-3", "Wk4+", "Baseline", "Wk2+"),
  inc_high = c(FALSE, FALSE, TRUE, FALSE, TRUE)
)

fnew(
  windows,
  type = "stratified_range",
  name = "visit_window",
  .other = "Outside window"
)

visit_df <- data.frame(
  arm = c("ARM_A", "ARM_A", "ARM_B", "ARM_B", "ARM_C"),
  day = c(3, 35, 5, 40, 10)
)
visit_df$window <- fputk(visit_df$arm, visit_df$day, format = "visit_window")
visit_df
```

This is useful for:

- visit window derivation in ADaM outputs
- protocol deviation checks
- windowed endpoint calculations

# 2. Composite keys for clinical mappings

Clinical trial data often uses multi-column logic: lab category + test code +
unit, or population + analysis flag + visit type. `fputk()` can build a lookup
from multiple fields without manually concatenating in the analysis step.

```{r}
#| label: composite-keys
fnew(
  fmap(
    paste(
      c("BLOOD CHEMISTRY", "COAGULOGRAM", "COAGULATION PANEL", "COAGULOGRAM"),
      c("BLOOD",           "BLOOD",        "BLOOD",             "BLOOD"),
      c("ALB",             "FIBRINO",      "INR",               "INR"),
      c("g/L",             "g/L",           NA,                  NA),
      sep = "|"
    ),
    c("ALB", "FIBRINO", "INR", "INR")
  ),
  type = "character",
  name = "lb_param",
  ignore_case = TRUE,
  .other = NA
)

lab_df <- data.frame(
  LBCAT = c("BLOOD CHEMISTRY", "COAGULOGRAM", "COAGULATION PANEL"),
  LBSPEC = c("BLOOD", "BLOOD", "BLOOD"),
  LBTESTCD = c("ALB", "INR", "INR"),
  LBSTRESU = c("g/L", NA, NA)
)

lab_df$PARAMCD <- fputk(
  lab_df$LBCAT,
  lab_df$LBSPEC,
  lab_df$LBTESTCD,
  lab_df$LBSTRESU,
  format = "lb_param",
  na_as_string = TRUE
)
lab_df
```

The same pattern is useful for:

- AE grouping by term + severity + relation
- CM coding by medication + route + dose form
- endpoint derivations from multiple source fields

# 3. Reverse lookup for QC and cleaning

`finputk()` is helpful when raw labels must be normalized back to codes. In
clinical programming, that makes it suitable for QC checks, mapping vendor text,
and validating reviewer-facing labels.

```{r}
#| label: reverse-lookup
finput(
  fmap(
    paste(c("PLACEBO", "DRUG A", "DRUG B"), c("PBO", "A50", "B100"), sep = "|"),
    c(1L, 2L, 3L)
  ),
  target_type = "integer",
  name = "trt_code_inv"
)

trt_df <- data.frame(
  ARMN = c("PLACEBO", "DRUG A", "DRUG B", "UNKNOWN"),
  TRTCD = c("PBO", "A50", "B100", "X")
)

trt_df$TRTSEQ <- finputk(trt_df$ARMN, trt_df$TRTCD, invalue_name = "trt_code_inv")
trt_df
```

Typical uses include:

- deriving internal codes from decoded labels in QC datasets
- validating data entry against controlled terminology
- reconciling spreadsheet exports back to source codes

# 4. Report-ready numeric formatting

`ksformat` can also be used where report output needs consistent numeric
presentation. This is not a lookup table; it is a pattern-driven formatter.

```{r}
#| label: numeric-display
fnew("$%,.2f", type = "numeric", name = "currency")
fnew("%.1f%%", type = "numeric", name = "pct")

summary_df <- data.frame(
  metric = c("Cost", "Response rate"),
  raw = c(1234.56, 0.153)
)

summary_df$display <- c(
  fputn(summary_df$raw[1], "currency"),
  fputn(summary_df$raw[2] * 100, "pct")
)
summary_df
```

This is useful for:

- CSR and TLF output generation
- QA listings for counts and rates
- ad hoc review tables where consistent display matters

# 5. Bucketing dates into operational periods

Clinical teams often work with periods such as screening, treatment, follow-up,
and analysis cut-offs. `date_range` and `datetime_range` let you keep those
rules in one place instead of duplicating them in ad hoc code.

```{r}
#| label: operational-periods
fparse(text = '
VALUE study_period (date_range)
  [2024-01-01, 2024-02-01) = "Screening"
  [2024-02-01, 2024-06-01) = "Treatment"
  [2024-06-01, HIGH]       = "Follow-up"
  .missing                 = "No date"
;
')

obs_dates <- as.Date(c("2024-01-15", "2024-03-10", "2024-07-01", NA))
data.frame(
  date = obs_dates,
  period = fput(obs_dates, "study_period")
)
```

# 6. Text generation for listings and narratives

Expression labels make `ksformat` useful for building small text fragments in
reports. In clinical settings, that can be handy for reviewer comments, summary
phrases, or footnotes.

```{r}
#| label: narrative-fragments
fnew(
  "high" = "sprintf('Above threshold: %s', .x1)",
  "low" = "sprintf('Below threshold: %s', .x1)",
  name = "thr_note"
)

fput(c("high", "low"), "thr_note", c(18.2, 7.4))
```

This is especially useful when the label depends on:

- a measurement value
- a threshold or cohort-specific cutoff
- an external text fragment such as a reviewer comment

# 7. Practical cautions

These patterns are powerful, but they work best when the mapping rules are
stable and owned by the analysis team. A few guardrails help:

- keep the format definition close to the analysis spec
- prefer a named format for rules reused across tables and datasets
- use `fclear()` in examples and tests to avoid leaking state between steps
- use `na_as_string = TRUE` only when the key was built with literal `"NA"`
  components from `paste()`

# Conclusion

`ksformat` is not limited to simple code-to-label dictionaries. In clinical
trials it can also serve as a compact rule engine for visit windows, composite
keys, reverse mappings, report formatting, and small narrative fragments.
That makes it useful anywhere repeated conditional logic would otherwise be
spread across scripts, spreadsheets, or one-off helper functions.

```{r}
#| label: cleanup
fclear()
```
